---
title: "Get started with SAVM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with SAVM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SAVM)
library(sf)
library(stars)
library(mapview)
```

# Introduction

This vignette provides an overview of the workflow offered by the **Submerged Aquatic Vegetation (SAV) Model R package**. The package allows users to import spatial and tabular data related to SAV presence and habitat conditions in aquatic ecosystems. The presented in this vignette provide an overview of the functionalities built into the package and presents a workflow from user input to model predictions. 


# User workflow

## Reading files

As an example, we consider a zone near Buffalo in the Lake Erie. Note that all the data 
required data are included on in the package SAVM.  

```{r}
# Lake Erie boundaries polygon
le_bound <- system.file("example", "lake_erie.gpkg", package = "SAVM") |>
      sf::st_read() |> 
      sf::st_transform(crs = 3857)

# Lake Erie study zone: read 
study_zone <- system.file("example", "study_zone.geojson", package = "SAVM") |>
  read_sav(spacing = 2000)

# Depth
study_depth <- stars::read_stars(system.file("example", "le_bathy.tiff", package = "SAVM"))

# Using mapview to map objects 
mapview(study_depth) + mapview(study_zone$polygon) + mapview(study_zone$points)
```


## Compute fetch and extract depth

### Copute fetch 

```{r fetch}
fetch <- compute_fetch(study_zone$points, le_bound, max_dist = 15)
fetch
```


### Extract depth 

```{r extract_depth}
depth <- stars::st_extract(study_depth, study_zone$points |> 
  sf::st_transform(crs = sf::st_crs(study_depth)))
```

Create an input data frame

```{r }
d_inputs <- data.frame(
  depth = depth$le_bathy.tiff,
  fetch = fetch$mean_fetch$fetch_km
)
```

## Model 

```{r}
# get input 
res <- sav_model(d_inputs)
res
```


## Vizualize

```{r}
# Visualize results 
library(ggplot2)
library(units)

study_zone$points$Cover <- res$Cover
study_zone$points$PA <- res$PA
study_zone$points$Depth <- res$Depth
study_zone$points$Fetch <- res$Fetch

plot_sav_distribution(study_zone$points  |> sf::st_drop_geometry())

plot_sav_density(study_zone$points |> sf::st_drop_geometry())
```


```{r}
mapview::mapview(study_zone$polygon, legend = FALSE, col.regions = "#7cdadc", alpha.regions = 0.2, layer.name = "Area of interest") + 
mapview::mapview(study_zone$points, zcol = "Cover", layer.name = "Cover", zlim = c(0,100)) +
mapview::mapview(study_zone$points, zcol = "PA", layer.name = "Presence", hide = TRUE) +
mapview::mapview(study_zone$points, zcol = "Depth", layer.name = "Depth", hide = TRUE) +
mapview::mapview(study_zone$points, zcol = "Fetch", layer.name = "Fetch", hide = TRUE)
```
